---
title: "Survival NPDR — Quick Demonstration"
author: "Kampmeyer, Dawkins, Urbanowicz, Kazemi-Nia, McKinney"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

This is a small, self-contained demonstration of Survival NPDR. It is designed
to run in minutes and does not reproduce the full paper benchmark. Use
`reproduce_full_paper.Rmd` for the complete workflow.

# 1 Setup

## Install packages (run once)

```{r install, eval=FALSE}
install.packages(c(
  "survival", "dplyr", "rlang", "PRROC", "glmnet",
  "caret", "survivalsvm", "ranger", "R.utils",
  "ggplot2", "tibble", "rsurv", "parallel",
  "survcomp", "broom", "here", "devtools"
))
```

## Install sNPDR (run once, from the repo root)

```{r install-snpdr, eval=FALSE}
devtools::install("../sNPDR")
```

## Load everything

```{r setup, message=FALSE, warning=FALSE}
library(survival)
library(dplyr)
library(ggplot2)
library(PRROC)
library(rsurv)
library(broom)

root <- here::here()   # repo root
devtools::load_all(file.path(root, "sNPDR"), quiet = TRUE)
```

---

# 2 Simulation Demo

A quick, self-contained run to verify the method end-to-end. We use smaller
parameters than the paper (p = 100, 3 replicates) so it finishes in a few
minutes on a laptop.

## Data generation

Survival times follow a lognormal AFT model:

$$\log T_i = \mu + \sum_{j=1}^{10} \beta_j X_{ij} + \sigma \varepsilon_i, \quad \varepsilon_i \sim \mathcal{N}(0,1)$$

The first 10 features are functional ($\beta \neq 0$); the remaining 90 are noise.

```{r simulate}
simulate_main <- function(n, p, n_main, beta, meanlog, sdlog, max_time, seed) {
  set.seed(seed)
  feat_names   <- paste0("simvar", seq_len(p))
  signal_names <- paste0("simvar", seq_len(n_main))
  X <- as.data.frame(matrix(rnorm(n * p), nrow = n, ncol = p,
                             dimnames = list(NULL, feat_names)))
  betas <- c(rep( beta, ceiling(n_main / 2)),
             rep(-beta, floor(n_main / 2)))
  fmla  <- as.formula(paste("~", paste(signal_names, collapse = " + ")))
  set.seed(seed + 1)
  t_event <- rsurv::raftreg(runif(n), formula = fmla, beta = betas,
                             dist = "lnorm", meanlog = meanlog,
                             sdlog = sdlog, data = X)
  time   <- pmin(t_event, max_time)
  status <- as.integer(t_event <= max_time)
  list(data = cbind(X, time = time, status = status),
       functional_vars = signal_names)
}

sim <- simulate_main(n = 200, p = 100, n_main = 10, beta = 0.4,
                     meanlog = log(100), sdlog = 0.25,
                     max_time = 125, seed = 2467)

cat(sprintf("n=%d | p=%d | events=%d (%.0f%%) | censored=%d\n",
            nrow(sim$data), ncol(sim$data) - 2,
            sum(sim$data$status), 100 * mean(sim$data$status),
            sum(sim$data$status == 0)))
```

## Run sNPDR

`diff.type = "signed"` preserves direction $(\Delta T \sim \Delta X)$ and is
optimal for main-effect detection in AFT models. `nbd.method = "multisurf"`
selects the neighborhood size adaptively.

```{r run-snpdr}
res <- npdr_surv(
  outcome    = c(time_var = "time", status_var = "status"),
  dataset    = sim$data,
  diff.type  = "signed",
  regression = "lm"
)

head(res[order(res$pval), ], 15)
```

## Enrichment check

```{r enrichment}
ranked <- res$feature[order(res$pval)]
func   <- sim$functional_vars

mrr_rank <- which(ranked %in% func)[1]
cat(sprintf("First functional feature at rank %d  (MRR = %.3f)\n",
            mrr_rank, 1 / mrr_rank))

k10  <- ceiling(100 * 0.10)   # top 10% of 100 features
hits <- sum(ranked[1:k10] %in% func)
ef10 <- hits / k10 / (10 / 100)
cat(sprintf("EF@10%% = %.1f  (%d / %d top features are functional)\n",
            ef10, hits, k10))
```

## auPRC across effect sizes (3-replicate mini sweep)

```{r mini-sweep}
beta_vals <- c(0.0, 0.2, 0.4)

auprc_df <- do.call(rbind, lapply(beta_vals, function(bv) {
  auprcs <- vapply(1:3, function(r) {
    s   <- simulate_main(200, 100, 10, bv, log(100), 0.25, 125, 2467 + r * 997L)
    out <- npdr_surv(
      outcome    = c(time_var = "time", status_var = "status"),
      dataset    = s$data,
      diff.type  = "signed",
      regression = "lm"
    )
    pos <- out$pval[ out$feature %in% s$functional_vars]
    neg <- out$pval[!out$feature %in% s$functional_vars]
    PRROC::pr.curve(scores.class0 = 1 - pos,
                    scores.class1 = 1 - neg)$auc.integral
  }, numeric(1))
  data.frame(beta = bv, mean_auprc = mean(auprcs), sd = sd(auprcs))
}))

ggplot(auprc_df, aes(x = beta, y = mean_auprc)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(size = 3, colour = "steelblue") +
  geom_errorbar(aes(ymin = mean_auprc - sd, ymax = mean_auprc + sd),
                width = 0.01, colour = "steelblue") +
  labs(x = expression(beta[main]), y = "auPRC",
       title = "sNPDR-signed: auPRC vs. effect size (demo: n=200, p=100, 3 reps)") +
  ylim(0, 1) + theme_bw()
```

---

# 3 Session Info

```{r session-info}
sessionInfo()
```
