---
title: "Survival NPDR — Full Paper Reproduction"
author: "Kampmeyer, Dawkins, Urbanowicz, Kazemi-Nia, McKinney"
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

This notebook is the full reproduction workflow for the paper. The simulation
chunks use the paper settings and can take hours. They are disabled by default;
set `eval = TRUE` for a chunk only when its prerequisites are installed and its
expected runtime is acceptable.

# 1 Setup

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)

root <- normalizePath(if (dir.exists("sNPDR")) "." else "..")
stopifnot(file.exists(file.path(root, "sNPDR", "DESCRIPTION")))
res_dir <- file.path(root, "results")
fig_dir <- file.path(root, "figures")
```

```{r install, eval=FALSE}
install.packages(c(
  "survival", "dplyr", "PRROC", "glmnet", "caret", "survivalsvm",
  "ranger", "R.utils", "ggplot2", "tibble", "rsurv", "parallel",
  "survcomp", "broom", "devtools"
))
devtools::install(file.path(root, "sNPDR"))
```

# 2 Simulation Benchmarks

The paper evaluates 20 replicated lognormal AFT simulations at 20% censoring.
Main-effect simulations use $n=200$, $p=1000$, and 10 causal attributes;
interaction simulations use $n=500$, $p=1000$, and 10 causal interaction pairs.

## Run the full sweeps

```{r run-effect-size-sweeps, eval=FALSE}
system2("Rscript", file.path(root, "scripts", "simulation",
  "main_effect_size_sweep_rsurv.r"))
system2("Rscript", file.path(root, "scripts", "simulation",
  "interaction_effect_size_sweep_rsurv.r"))
```

## Figures 2 and 3 — Main and interaction effects

```{r plot-effect-size-figures, eval=FALSE}
system2("Rscript", file.path(root, "scripts", "figures",
  "plot_ef_vs_effect_size.R"))
```

```{r load-effect-size-results}
main_summary <- read.csv(file.path(res_dir,
  "main_effect_size_sweep_rsurv_results_summary.csv"))
int_summary <- read.csv(file.path(res_dir,
  "interaction_effect_size_sweep_rsurv_results_summary.csv"))

method_colours <- c(
  "Cox" = "#E41A1C", "SVM" = "#FF7F00", "LASSO-Cox" = "#F781BF",
  "sNPDR-signed" = "#1F78B4", "sNPDR-absolute" = "#984EA3",
  "LASSO-sNPDR" = "#A65628", "Ranger" = "#4DAF4A"
)
```

```{r figure-2}
main_summary %>%
  filter(method %in% names(method_colours)) %>%
  mutate(method = factor(method, levels = names(method_colours))) %>%
  ggplot(aes(beta_main, mean_ef_at_1pct, colour = method, group = method)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = mean_ef_at_1pct - sd_ef_at_1pct,
                    ymax = mean_ef_at_1pct + sd_ef_at_1pct), width = 0.008) +
  scale_colour_manual(values = method_colours, name = NULL) +
  labs(x = expression(beta[main]), y = "Enrichment Factor @ 1%",
       title = "Figure 2. Main-effect simulation performance") +
  theme_bw()
```

```{r figure-3}
int_summary %>%
  filter(method %in% names(method_colours)) %>%
  mutate(method = factor(method, levels = names(method_colours))) %>%
  ggplot(aes(beta_int, mean_ef_at_1pct, colour = method, group = method)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = mean_ef_at_1pct - sd_ef_at_1pct,
                    ymax = mean_ef_at_1pct + sd_ef_at_1pct), width = 0.004) +
  scale_colour_manual(values = method_colours, name = NULL) +
  labs(x = expression(beta[int]), y = "Enrichment Factor @ 1%",
       title = "Figure 3. Interaction-effect simulation performance") +
  theme_bw()
```

## Tables 1 and 2 — Paper effect sizes

```{r table-1}
main_summary %>%
  filter(beta_main == 0.13) %>%
  select(method, mean_auprc, mean_mrr, mean_ef_at_1pct, mean_ef_at_5pct,
         mean_ef_at_10pct, mean_runtime, n_sims) %>%
  arrange(desc(mean_ef_at_1pct)) %>%
  kable(digits = 3, caption = "Table 1. Main-effect simulation at beta = 0.13.")
```

```{r table-2}
int_summary %>%
  filter(beta_int == 0.20) %>%
  select(method, mean_auprc, mean_mrr, mean_ef_at_1pct, mean_runtime, n_sims) %>%
  arrange(desc(mean_ef_at_1pct)) %>%
  kable(digits = 3, caption = "Table 2. Interaction-effect simulation at beta = 0.20.")
```

# 3 Figure 4 — Neighborhood Size Sensitivity

The main-effect sweep uses $p=1000$. The interaction sweep uses $p=100$.

```{r run-k-sweep, eval=FALSE}
system2("Rscript", c(file.path(root, "scripts", "test_k", "test_k_lognormal.R"),
                      "0.5", "1000", "100"))
system2("Rscript", c(file.path(root, "scripts", "test_k",
                                "plot_k_combined_lognormal.R"), "0.5", "100"))
```

```{r figure-4, eval=FALSE}
knitr::include_graphics(file.path(fig_dir,
  "auPRC_K_combined_lognormal_no_multisurf_sigma0.50_p100.png"))
```

# 4 Real Data

## Figure 5 and Table 3 — Breast cancer

The raw breast-expression data are large and are not versioned in Git. Download
and preprocess them before running these chunks. The paper workflow uses the
top 5000 genes and defines high- and low-risk groups as the upper and lower
40% of predicted Cox risk scores.

```{r prepare-breast-data, eval=FALSE}
system2("Rscript", file.path(root, "scripts", "real_data",
  "preprocess_breast_cancer.R"))
system2("Rscript", file.path(root, "scripts", "real_data",
  "breast_model_performance.R"))
system2("Rscript", file.path(root, "scripts", "real_data",
  "breast_survival_curve.R"))
```

## Table 4 — Ovarian cancer

```{r ovarian-benchmark, eval=FALSE}
system2("Rscript", file.path(root, "scripts", "real_data",
  "ovarian_risk_stratification.R"))
```

# 5 Figures 6 and 7 — Interaction Networks

Figure 7 is generated by the ovarian ReGAIN workflow below. The breast-network
source workflow required for Figure 6 is not currently included in this
repository; therefore this notebook cannot regenerate Figure 6 until that
script and its input data are restored.

```{r ovarian-network, eval=FALSE}
system2("Rscript", file.path(root, "scripts", "network",
  "ovarian_cancer_network.r"))
```

# 6 Statistical Analysis

```{r statistical-tests, eval=FALSE}
system2("Rscript", file.path(root, "scripts", "analysis_statistical.R"))
```

# Session Information

```{r session-info}
sessionInfo()
```